---
title: "FinalProject"
output: html_document
date: "2026-02-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(forecast)
library(tseries)
library(ggplot2)
library(fpp3)
```

# 1. Data Loading

```{r initial_analysis, echo=TRUE}
# Read in data
data <- read.csv("IMPCH.csv")
impch <- data$IMPCH

# Create ts object (yearly)
impch_ts <- ts(impch, start=c(1985,1), frequency=12)

# Check for missing values
sum(is.na(impch_ts)) # result here is 0

```

# 2. Preprocessing
We can do some analysis here on trend / seasonality/ whatever. We can also split our training data now.
```{r pressure, echo=TRUE}
# TODO: normalize dataset, will need to in all likelihood. Run tests for all those things

# Observe raw series
autoplot(impch_ts) + ggtitle("IMPCH (level)")
```
The plot shows clear strong upward trend and increasing variability (variance grows as values get larger), therefore will need a Box-Cox transformation.

```{r pressure, echo=TRUE}
# Find optimal lambda
lambda <- BoxCox.lambda(impch_ts) 
impch_bc <- BoxCox(impch_ts, lambda)

autoplot(impch_bc) + ggtitle("log(IMPCH (level))")
```
With lambda = 0.102, variance has stabilized significantly, with less wild swings at the end of the timeframe.

```{r pressure, echo=TRUE}
# Train & test split for time series
train <- window(impch_bc, end = c(2020, 12))
test <- window(impch_bc, start = c(2021, 1))

# Verify split
cat("TRAIN: Jan 1985 to Dec 2020 (", length(train), "obs)\n")
cat("TEST:  Jan 2021 to Dec 2025 (", length(test),  "obs)\n")
cat("Train %: ", round(100 * length(train) / length(impch_ts), 1), "%\n")
```
We used a defined train/test split (vs. the conventional 80/20) to incorporate COVID-19 effects while maintaining a meaningful out-of-sample test period, for two reasons:

1. COVID-aware training: Includes 2020 to capture the pandemic onset and initial economic shock, allowing whatever model we use to learn from this structural break.

2. Realistic forecasting validation: The 2021-2025 test period covers post-COVID recovery and stabilization, and provides a realistic test of forecast performance under the 'new normal' conditions.


# Model Fitting
```{r model_fit, echo=TRUE}
# Fit at least three different types of models for evaluation. We can choose one of the sliding windows to use here as well
```

### Model #1: ARIMA
```{r model_fit, echo=TRUE}
# Non-seasoning differencing
diff1 <- diff(train)
diff2 <- diff(train, 2)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

Acf(diff1, main = "ACF of 1st difference")
Pacf(diff1, main = "PACF of 1st difference")
Acf(diff2, main = "ACF of 2nd difference")
Pacf(diff2, main = "PACF of 2nd difference")

par(mfrow = c(1, 1))
```
ACF/PACF of 1st difference shows lags mostly within the bands, expect a few seasonal spikes. ACF/PACF of 2nd difference looks much noisier, with larger spikes, indicating potentially over-differencing.

```{r model_fit, echo=TRUE}
# Seasonal differencing
diff_seasonal <- diff(train, lag = 12)
diff_both <- diff(diff(train), lag = 12)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

Acf(diff_seasonal, main = "ACF of D = 1", lag.max = 36)
Pacf(diff_seasonal, main = "PACF of D = 1", lag.max = 36)
Acf(diff_both, main = "ACF of d = 1, D = 1", lag.max = 36)
Pacf(diff_both, main = "PACF of d = 1, D = 1", lag.max = 36)
```
ACF after seasonal differencing only (D = 1) shows a slow decay across lags, indicating that the series is still non-stationary and seaosnal differencing alone is not enough.

ACF after both differencing (d = 1, D = 1), shows autocorrelations that are now mostly small and within the blue bounds, except maybe one seaosnal spike around lag 12.

Two potential SARIMA candidates to investigate are:
1. (1,1,0)(0,1,1) - one noticeable spike at lag 1 for PACF (likely AR(1)), one noticeable spike at lag 1 for ACF (likely MA(1))
2. (0,1,1)(0,1,1) - ACF spike at lag 12 (likely MA(1))

```{r model_fit, echo=TRUE}
# Model 1: ARIMA(1,1,0)(0,1,1)
arima_fit1 <- Arima(train,
              order = c(1,1,0),
              seasonal = list(order = c(0,1,1), period = 12))

# Model 2: ARIMA(0,1,1)(0,1,1)
arima_fit2 <- Arima(train,
              order   = c(0,1,1),
              seasonal = list(order = c(0,1,1), period = 12))

# Model 3: Using auto.arima to choose best model
arima_fit3 <- auto.arima(train, 
                   seasonal = TRUE,
                   stepwise = FALSE,
                   approximation = FALSE)

model_compare <- data.frame(
  Model = c("ARIMA(1,1,0)(0,1,1)",
            "ARIMA(0,1,1)(0,1,1)",
            "auto.arima"),
  AICc = round(c(arima_fit1$aicc, arima_fit2$aicc, arima_fit3$aicc), 2),
  BIC = round(c(BIC(arima_fit1),  BIC(arima_fit2),  BIC(arima_fit3)),  2)
)
model_compare
print (arima_fit3)
```
The auto.arima model ARIMA(2,1,1)(2,1,0) has lower AICc and BIC, indicating the model that provides the best balance between fitting the data well and keeping the model simple / avoiding overfitting.

# Model Forecasting
Do the extra stuff here

### Model #1: ARIMA
```{r forecast, echo=TRUE}
best_arima <- arima_fit3

h <- length(test)
fc_arima <- forecast(best_arima, h = h)

fc_impch_arima <- fc_arima
fc_impch_arima$mean <- InvBoxCox(fc_arima$mean,  lambda)
fc_impch_arima$lower <- InvBoxCox(fc_arima$lower, lambda)
fc_impch_arima$upper <- InvBoxCox(fc_arima$upper, lambda)
fc_impch_arima$x <- impch_ts

autoplot(fc_impch_arima) +
  ggtitle("IMPCH: ARIMA forecasts for next periods") +
  ylab("IMPCH") +
  xlab("Time")
```
Based on below residuals for best_arima, it looks like although residuals are normally distributed and resemble white noise with mean zero, our ACF still shows significant lags (7 and 11)
```{r forecast, echo=TRUE}
checkresiduals(best_arima)

fc_all <- forecast(best_arima, h = h)
fc_all_bc <- fc_all
fc_all_bc$mean <- InvBoxCox(fc_all$mean, lambda)

fc_ts <- ts(InvBoxCox(fc_all$mean, lambda),
            start     = start(test),
            frequency = frequency(test))

autoplot(impch_ts) +
  autolayer(fc_ts, series = "ARIMA forecast") +
  ggtitle("IMPCH: auto.arima forecast vs full series") +
  labs(x = "Time", y = "IMPCH")
```

## Model 2 (VAR)


## Model 3 (Sine / Cosine)

```{r fourier, echo=TRUE}
# start with spectral analysis
library(TSA)
library
df_periodogram <- periodogram(diff1)
dd <- data.frame(freq = df_periodogram$freq, spec = df_periodogram$spec)
periodogram_sorted <- dd[order(-dd$spec), ]
head(periodogram_sorted)
```
We see that the highest frequencies are .083 (represents annual seasonality), (perhaps quarterly, since its roughly every 3 months), .25 (every four months)

```{r model_expand_slide}